---
title: "Forest plots"
subtitle: "Source notebook for figures 2–4"
author: "PTCy meta-analysis collaboration"
format:
  html:
    toc: true
    code-fold: show
    fig-width: 8
    fig-height: 9
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---

The figures in the article are drawn from two ingredients that come from different
places:

- **the study rows** — raw per-study odds ratios with Wald 95% CIs, computed here
  directly from the extracted event counts in `data/analytic/`, with a 0·5 continuity
  correction applied only to studies with a zero cell;
- **the pooled row** — the Bayesian random-effects posterior (model M1), taken from the
  exported summaries in `data/models/Table2_setB.csv`.

The study rows are *not* per-study Bayesian estimates. M1 has a single shared treatment
effect with a random intercept for study-level baseline risk only, so individual studies
have no treatment-effect posterior of their own.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(stringr)

here <- function(...) file.path(if (basename(getwd()) == "notebooks") ".." else ".", ...)

studies <- read_csv(here("data", "studies.csv"), show_col_types = FALSE) |>
  select(study_id, first_author, pub_year)
pooled  <- read_csv(here("data", "models", "Table2_setB.csv"), show_col_types = FALSE)

# ── Lancet Haematology palette ──
# lancet_red is sampled from the journal's own PDF; the greys are the neutrals
# it sets rules and secondary text in.
lancet_red  <- "#b20d35"
lancet_dark <- "#2d2d2d"
lancet_grey <- "#8c8c8c"


# The journal draws forest plots on a clean white field: no grid, a single
# baseline rule, titles in the sans face, everything else out of the way.
theme_forest <- function(base_size = 10) {
  theme_minimal(base_size = base_size) +
    theme(panel.grid        = element_blank(),
          axis.line.x  = element_line(colour = lancet_dark, linewidth = .4),
          axis.ticks.x = element_line(colour = lancet_dark, linewidth = .3),
          axis.ticks.length.x = unit(2.5, "pt"),
          axis.text    = element_text(colour = lancet_dark),
          axis.title.x = element_text(colour = lancet_dark, size = rel(.9),
                                      margin = margin(t = 6)),
          plot.title    = element_text(face = "bold", colour = lancet_dark,
                                       hjust = 0, size = rel(1.05)),
          plot.subtitle = element_text(colour = lancet_grey, hjust = 0, size = rel(.85)),
          plot.caption  = element_text(colour = lancet_grey, hjust = 0, size = rel(.75)),
          plot.title.position = "plot",
          plot.caption.position = "plot",
          plot.background  = element_rect(fill = "white", colour = NA),
          panel.background = element_rect(fill = "white", colour = NA))
}
```

```{r}
#| label: functions

# Wald log-OR and CI from a 2x2 table, with 0.5 added only where a cell is zero.
study_or <- function(d) {
  d |>
    mutate(zero = ptcy_e == 0 | comp_e == 0 |
                  ptcy_e == ptcy_n | comp_e == comp_n,
           a = ptcy_e + 0.5 * zero, b = ptcy_n - ptcy_e + 0.5 * zero,
           c = comp_e + 0.5 * zero, d2 = comp_n - comp_e + 0.5 * zero,
           yi = log((a / b) / (c / d2)),
           sei = sqrt(1/a + 1/b + 1/c + 1/d2),
           or = exp(yi), lo = exp(yi - 1.96 * sei), hi = exp(yi + 1.96 * sei)) |>
    filter(is.finite(or), is.finite(lo), is.finite(hi))
}

# The adopted analysis is Set B: one publication per cohort per outcome. Only the
# outcomes affected by deduplication were refitted, so data/models/ is a partial overlay
# -- prefer the Set B dataset for a slug, and fall back to the full analytic dataset for
# the outcomes deduplication left untouched.
resolve_data <- function(analytic_file, slug) {
  setb <- here("data", "models", paste0("data_", slug, ".csv"))
  if (file.exists(setb)) return(setb)
  here("data", "analytic", analytic_file)
}

forest <- function(analytic_file, slug, title, subtitle = NULL) {
  f <- resolve_data(analytic_file, slug)
  if (!file.exists(f)) { message("missing: ", analytic_file); return(invisible(NULL)) }
  message("  ", slug, " <- ", sub("^.*/", "", f))

  es <- read_csv(f, show_col_types = FALSE) |>
    study_or() |>
    left_join(studies, by = "study_id") |>
    mutate(slab = sprintf("%s (%s)", first_author, pub_year))
  # multi-arm studies contribute more than one row: keep the labels distinct
  es$slab <- make.unique(es$slab, sep = " – ")
  es <- es |> arrange(yi) |> mutate(slab = factor(slab, levels = slab))

  p <- pooled |> filter(slug == !!slug, model == "m1")
  stopifnot(nrow(p) == 1)

  pd <- bind_rows(
    es |> transmute(slab = as.character(slab), or, lo, hi, pooled = FALSE),
    tibble(slab = "Bayesian pooled (M1)", or = p$or_median,
           lo = p$ci_low, hi = p$ci_high, pooled = TRUE))
  pd$slab <- factor(pd$slab, levels = c("Bayesian pooled (M1)", levels(es$slab)))

  ggplot(pd, aes(or, slab)) +
    geom_vline(xintercept = 1, linetype = "dashed",
               colour = lancet_grey, linewidth = .35) +
    # study rows: crimson squares, the journal's forest-plot marker
    geom_pointrange(data = filter(pd, !pooled), aes(xmin = lo, xmax = hi),
                    shape = 15, size = .3, linewidth = .4, colour = lancet_red) +
    # pooled row: a filled diamond, larger and in the same crimson
    geom_pointrange(data = filter(pd, pooled), aes(xmin = lo, xmax = hi),
                    shape = 18, size = .95, linewidth = .8, colour = lancet_red) +
    scale_x_log10() +
    labs(x = "Odds ratio (log scale)", y = NULL, title = title,
         subtitle = subtitle %||% sprintf("k = %d, N = %s, τ = %.2f",
                                          p$k, format(p$n_total, big.mark = " "), p$tau),
         caption = "Study rows: raw per-study OR with Wald 95% CI. Diamond: Bayesian posterior median with 95% CrI.") +
    theme_forest() +
    theme(axis.text.y = element_text(
      face = ifelse(levels(pd$slab) == "Bayesian pooled (M1)", "bold", "plain"),
      size = rel(.85)))
}

`%||%` <- function(a, b) if (is.null(a)) b else a
```

## Figure 2 — overall survival (comparison 1)

```{r}
#| label: fig-os
#| fig-cap: "Forest plot for overall survival, comparison 1 (PTCy vs CNI + MTX/MMF)."
#| fig-height: 10
forest("C1_overall_mortality_OS_event.csv", "c1_os",
       "Overall survival — PTCy vs CNI + MTX/MMF")
```

## Figure 3 — CMV reactivation

```{r}
#| label: fig-cmv-c1
#| fig-cap: "CMV any-reactivation, comparison 1 (PTCy vs CNI + MTX/MMF)."
#| fig-height: 7
forest("C1_CMV_any_reactivation.csv", "c1_cmv",
       "CMV reactivation — PTCy vs CNI + MTX/MMF")
```

```{r}
#| label: fig-cmv-c2
#| fig-cap: "CMV any-reactivation, comparison 2 (PTCy vs ATG). The direction reverses relative to comparison 1."
#| fig-height: 5
forest("C2_CMV_any_reactivation.csv", "c2_cmv",
       "CMV reactivation — PTCy vs ATG")
```

## Figure 4 — bacterial, fungal, and BK viral infection

```{r}
#| label: fig-bsi
#| fig-cap: "Bloodstream infection, any pathogen (comparison 1)."
#| fig-height: 4
forest("C1_BSI_any_pathogen.csv", "c1_bsi",
       "Bloodstream infection, any pathogen — PTCy vs CNI + MTX/MMF")
```

```{r}
#| label: fig-ifi
#| fig-cap: "Invasive fungal infection, any (comparison 1). Note the very large τ."
#| fig-height: 4
forest("C1_IFI_any_EORTC_MSG_proven_probable_combined.csv", "c1_ifi_any",
       "Invasive fungal infection, any — PTCy vs CNI + MTX/MMF")
```

Panel C of figure 4 in the article is BK virus reactivation. The BK analytic dataset was
not part of the `02_extraction/analytic/` export in the analysis repository, so it is not
reproduced here — the pooled estimate is still carried in table 2
(`c1_bk`, k = 11, OR 2·48 [1·82–3·38]). To add the panel, export the BK pairs from
`outcomes.csv` into `data/analytic/C1_BK_virus_reactivation.csv` with the same columns
as the other files and call `forest()` on it.

## Acute GVHD grade II–IV

```{r}
#| label: fig-agvhd
#| fig-cap: "Acute GVHD grade II–IV (comparison 1)."
#| fig-height: 8
forest("C1_aGVHD_grade_II_IV.csv", "c1_agvhd",
       "Acute GVHD grade II–IV — PTCy vs CNI + MTX/MMF")
```

## Exporting the article figures

Two of the article's figures changed when the manuscript moved to Set B: overall
survival (comparison 1) and the comparison 2 panel of the CMV figure. Both are exported
here in all three renditions so that `index.qmd` picks up the right one per output
format, and so that the figures cannot disagree with table 2.

The remaining article figures — the comparison 1 CMV panel, and the whole of figure 4
(BSI, IFI, BK) — were unaffected by deduplication, so the originals from the analysis
repository are kept.

```{r}
#| label: export-figures
#| output: false

export <- function(plot, stem, w = 7, h = 9) {
  for (dev in c("png", "pdf", "svg")) {
    ggsave(here("figures", paste0(stem, ".", dev)), plot,
           width = w, height = h, dpi = 300,
           device = dev, bg = "white")
  }
  invisible(stem)
}

export(forest("C1_overall_mortality_OS_event.csv", "c1_os",
              "Overall survival — PTCy vs CNI + MTX/MMF"),
       "Figure2_OS_forest_setB", w = 7, h = 9)

export(forest("C2_CMV_any_reactivation.csv", "c2_cmv",
              "CMV reactivation — PTCy vs ATG"),
       "Figure3b_CMV_C2_forest_setB", w = 7, h = 4.5)

# Panel A of the CMV figure is unchanged by deduplication, but the repository ships it
# baked into the same two-panel image as panel B, so it is redrawn here to keep both
# panels on one source and one visual style.
export(forest("C1_CMV_any_reactivation.csv", "c1_cmv",
              "CMV reactivation — PTCy vs CNI + MTX/MMF"),
       "Figure3a_CMV_C1_forest_setB", w = 7, h = 6.5)
```

## Available analytic datasets

```{r}
#| label: tbl-analytic
tibble(file = list.files(here("data", "analytic"), pattern = "[.]csv$")) |>
  filter(!str_starts(file, "_")) |>
  rowwise() |>
  mutate(k = nrow(read_csv(here("data", "analytic", file), show_col_types = FALSE))) |>
  ungroup() |>
  knitr::kable()
```
