---
title: "Forest plots"
subtitle: "Source notebook for figures 2–4"
author: "PTCy meta-analysis collaboration"
format:
  html:
    toc: true
    code-fold: show
    fig-width: 9
    fig-height: 9
    fig-dpi: 300
execute:
  echo: true
  warning: false
  message: false
---

The figures in the article are drawn from two ingredients that come from different
places:

- **the study rows** — raw per-study odds ratios with Wald 95% CIs, computed here
  directly from the extracted event counts, with a 0·5 continuity correction applied
  only to studies with a zero cell;
- **the pooled row** — the Bayesian random-effects posterior (model M1), taken from the
  exported summaries in `data/models/Table2_setB.csv`.

The study rows are *not* per-study Bayesian estimates. M1 carries a random study
intercept and a random PTCy-effect slope, so each study does have a (shrunken)
study-specific effect posterior; the rows shown here are deliberately the raw,
unshrunken per-study estimates, with pooling left to the Bayesian diamond.

Every plot shows, beside the point estimates, the events and totals of each arm and
the REML inverse-variance weight of each study, as the journal's forest plots require.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(stringr)
library(metafor)
library(patchwork)
library(ggtext)

here <- function(...) file.path(if (basename(getwd()) == "notebooks") ".." else ".", ...)

studies <- read_csv(here("data", "studies.csv"), show_col_types = FALSE) |>
  select(study_id, first_author, pub_year)
pooled  <- read_csv(here("data", "models", "Table2_setB.csv"), show_col_types = FALSE)

# ── Lancet Haematology palette ──
lancet_red  <- "#b20d35"
lancet_dark <- "#2d2d2d"
lancet_grey <- "#8c8c8c"

theme_forest <- function(base_size = 10) {
  theme_minimal(base_size = base_size) +
    theme(panel.grid        = element_blank(),
          axis.line.x  = element_line(colour = lancet_dark, linewidth = .4),
          axis.ticks.x = element_line(colour = lancet_dark, linewidth = .3),
          axis.ticks.length.x = unit(2.5, "pt"),
          axis.text    = element_text(colour = lancet_dark),
          axis.title.x = element_text(colour = lancet_dark, size = rel(.9),
                                      margin = margin(t = 6)),
          plot.title    = element_text(face = "bold", colour = lancet_dark,
                                       hjust = 0, size = rel(1.05)),
          plot.subtitle = element_text(colour = lancet_grey, hjust = 0, size = rel(.85)),
          plot.caption  = element_text(colour = lancet_grey, hjust = 0, size = rel(.75)),
          plot.title.position = "plot",
          plot.caption.position = "plot",
          plot.background  = element_rect(fill = "white", colour = NA),
          panel.background = element_rect(fill = "white", colour = NA))
}
```

```{r}
#| label: functions

# Shared forest-panel implementation (single source of truth, also used by the
# appendix figure S8 series). Pass the project `here()` so paths resolve from
# either the project root or the notebooks/ directory.
source(here("scripts", "forest-lib.R"))

`%||%` <- function(a, b) if (is.null(a)) b else a
```

```{r}
#| label: functions-legacy
#| eval: false

# Wald log-OR and CI from a 2x2 table, with 0.5 added only where a cell is zero.
study_or <- function(d) {
  d |>
    mutate(zero = ptcy_e == 0 | comp_e == 0 |
                  ptcy_e == ptcy_n | comp_e == comp_n,
           a = ptcy_e + 0.5 * zero, b = ptcy_n - ptcy_e + 0.5 * zero,
           c = comp_e + 0.5 * zero, d2 = comp_n - comp_e + 0.5 * zero,
           yi = log((a / b) / (c / d2)),
           sei = sqrt(1/a + 1/b + 1/c + 1/d2),
           or = exp(yi), lo = exp(yi - 1.96 * sei), hi = exp(yi + 1.96 * sei)) |>
    filter(is.finite(or), is.finite(lo), is.finite(hi))
}

# The adopted analysis is Set B: one publication per cohort per outcome. Only the
# outcomes affected by deduplication were refitted, so data/models/ is a partial overlay
# -- prefer the Set B dataset for a slug, and fall back to the full analytic dataset for
# the outcomes deduplication left untouched.
resolve_data <- function(analytic_file, slug) {
  setb <- here("data", "models", paste0("data_", slug, ".csv"))
  if (file.exists(setb)) return(setb)
  here("data", "analytic", analytic_file)
}

POOLED <- "Bayesian pooled (M1)"
HEADER <- "column_header"

# One forest panel: point estimates on the left, events/totals and REML weights
# in a side table on the right. The pooled row sits at the top and is the only
# bold row.
forest <- function(analytic_file, slug, title, subtitle = NULL) {
  f <- resolve_data(analytic_file, slug)
  if (!file.exists(f)) { message("missing: ", analytic_file); return(invisible(NULL)) }
  message("  ", slug, " <- ", sub("^.*/", "", f))

  es <- read_csv(f, show_col_types = FALSE) |>
    study_or() |>
    left_join(studies, by = "study_id") |>
    mutate(slab = sprintf("%s (%s)", first_author, pub_year))
  # multi-arm studies contribute more than one row: keep the labels distinct
  es$slab <- make.unique(es$slab, sep = " – ")

  # REML inverse-variance weights (the frequentist concordance convention)
  es$wt <- as.numeric(weights(rma(yi, sei = sei, data = es, method = "REML",
                                  control = list(silent = TRUE))))

  p <- pooled |> filter(slug == !!slug, model == "m1")
  stopifnot(nrow(p) == 1)

  pd <- bind_rows(
    es |>
      arrange(yi) |>
      transmute(slab, or, lo, hi,
                ev_p = sprintf("%d/%d", ptcy_e, ptcy_n),
                ev_c = sprintf("%d/%d", comp_e, comp_n),
                wt = sprintf("%.1f", wt), pooled = FALSE),
    tibble(slab = POOLED, or = p$or_median, lo = p$ci_low, hi = p$ci_high,
           ev_p = "—", ev_c = "—", wt = "—", pooled = TRUE)
  )
  hdr <- tibble(slab = HEADER, or = NA_real_, lo = NA_real_, hi = NA_real_,
                ev_p = "PTCy events/N", ev_c = "Comparator events/N",
                wt = "Weight (%)", pooled = FALSE)
  pd <- bind_rows(pd, hdr) |>
    mutate(slab = factor(slab, levels = c(es$slab[order(es$yi)], POOLED, HEADER)))

  main <- ggplot(pd, aes(or, slab)) +
    geom_vline(xintercept = 1, linetype = "dashed",
               colour = lancet_grey, linewidth = .35) +
    geom_pointrange(data = filter(pd, !pooled & slab != HEADER),
                    aes(xmin = lo, xmax = hi),
                    shape = 15, size = .3, linewidth = .4, colour = lancet_red) +
    geom_pointrange(data = filter(pd, pooled),
                    aes(xmin = lo, xmax = hi),
                    shape = 18, size = .95, linewidth = .8, colour = lancet_red) +
    # No bold rows anywhere: vectorised face= is unsupported and misassigns
    # weights, and markdown labels did not parse reliably here. The pooled row
    # is distinguished by the diamond and its top position; the header row's
    # label is blanked.
    scale_x_log10() +
    scale_y_discrete(labels = \(x) ifelse(x == HEADER, "", x)) +
    labs(x = "Odds ratio (log scale)", y = NULL, title = title,
         subtitle = subtitle %||% sprintf("k = %d, N = %s, τ = %.2f",
                                          p$k, format(p$n_total, big.mark = " "), p$tau),
         caption = paste0("Study rows: raw per-study OR with Wald 95% CI (0·5 correction for zero cells). ",
                          "Weights: REML inverse-variance. Diamond: Bayesian posterior median, 95% CrI (M1).")) +
    theme_forest() +
    theme(axis.text.y = element_text(size = rel(.85)))

  side <- ggplot(pd, aes(y = slab)) +
    geom_text(aes(x = 1, label = ev_p), size = 2.4, colour = lancet_dark) +
    geom_text(aes(x = 2, label = ev_c), size = 2.4, colour = lancet_dark) +
    geom_text(aes(x = 3, label = wt), size = 2.4, colour = lancet_dark) +
    scale_x_continuous(limits = c(0.4, 3.6), expand = c(0, 0)) +
    scale_y_discrete(labels = \(x) ifelse(x == HEADER, "", x)) +
    labs(x = NULL, y = NULL) +
    theme_forest() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.ticks.length = unit(0, "pt"),
          axis.line.x = element_blank(),
          axis.text.y = element_text(size = rel(.85)),
          plot.caption = element_blank())

  main + side + plot_layout(widths = c(2.5, 1.9))
}

`%||%` <- function(a, b) if (is.null(a)) b else a
```

## Figure 2 — overall survival (comparison 1)

```{r}
#| label: fig-os
#| fig-cap: "Forest plot for overall survival, comparison 1 (PTCy vs CNI + MTX/MMF)."
#| fig-height: 10
forest("C1_overall_mortality_OS_event.csv", "c1_os",
       "Overall survival — PTCy vs CNI + MTX/MMF")
```

## Figure 3 — CMV reactivation

```{r}
#| label: fig-cmv-c1
#| fig-cap: "CMV any-reactivation, comparison 1 (PTCy vs CNI + MTX/MMF)."
#| fig-height: 7.5
forest("C1_CMV_any_reactivation.csv", "c1_cmv",
       "CMV reactivation — PTCy vs CNI + MTX/MMF")
```

```{r}
#| label: fig-cmv-c2
#| fig-cap: "CMV any-reactivation, comparison 2 (PTCy vs ATG). The direction reverses relative to comparison 1."
#| fig-height: 5.5
forest("C2_CMV_any_reactivation.csv", "c2_cmv",
       "CMV reactivation — PTCy vs ATG")
```

## Figure 4 — bacterial, fungal, and BK viral infection

The three panels are drawn from the same model-input datasets that feed table 2
(vendored into `data/models/` from the analysis repository), so the study rows and
the pooled diamond cannot disagree.

```{r}
#| label: fig-infections-panels
#| fig-cap: "Bloodstream infection, invasive fungal infection, and BK virus (comparison 1)."
#| fig-height: 9
p_bsi <- forest("C1_BSI_any_pathogen.csv", "c1_bsi",
                "Bloodstream infection, any pathogen — PTCy vs CNI + MTX/MMF")
p_ifi <- forest("C1_IFI_any_EORTC_MSG_proven_probable_combined.csv", "c1_ifi_any",
                "Invasive fungal infection, any — PTCy vs CNI + MTX/MMF")
p_bk  <- forest("C1_BK_virus_reactivation.csv", "c1_bk",
                "BK virus reactivation — PTCy vs CNI + MTX/MMF")
p_bsi / p_ifi / p_bk
```

## Acute GVHD grade II–IV

```{r}
#| label: fig-agvhd
#| fig-cap: "Acute GVHD grade II–IV (comparison 1)."
#| fig-height: 8.5
forest("C1_aGVHD_grade_II_IV.csv", "c1_agvhd",
       "Acute GVHD grade II–IV — PTCy vs CNI + MTX/MMF")
```

## Exporting the article figures

All article figures are exported here in all three renditions (PNG at 300 dpi for
the web build, vector PDF and SVG for the journal), so that `index.qmd` picks up the
right one per output format, and so that the figures cannot disagree with table 2.

```{r}
#| label: export-figures
#| output: false

export <- function(plot, stem, w = 9, h = 9) {
  for (dev in c("png", "pdf", "svg")) {
    ggsave(here("figures", paste0(stem, ".", dev)), plot,
           width = w, height = h, dpi = 300,
           device = dev, bg = "white")
  }
  invisible(stem)
}

export(forest("C1_overall_mortality_OS_event.csv", "c1_os",
              "Overall survival — PTCy vs CNI + MTX/MMF"),
       "Figure2_OS_forest_setB", w = 9, h = 9.5)

export(forest("C1_CMV_any_reactivation.csv", "c1_cmv",
              "CMV reactivation — PTCy vs CNI + MTX/MMF"),
       "Figure3a_CMV_C1_forest_setB", w = 9, h = 7)

export(forest("C2_CMV_any_reactivation.csv", "c2_cmv",
              "CMV reactivation — PTCy vs ATG"),
       "Figure3b_CMV_C2_forest_setB", w = 9, h = 5.5)

fig4 <- p_bsi / p_ifi / p_bk
export(fig4, "Figure4_infections_forest", w = 9, h = 9)
```

## Available analytic datasets

```{r}
#| label: tbl-analytic
tibble(file = list.files(here("data", "analytic"), pattern = "[.]csv$")) |>
  filter(!str_starts(file, "_")) |>
  rowwise() |>
  mutate(k = nrow(read_csv(here("data", "analytic", file), show_col_types = FALSE))) |>
  ungroup() |>
  knitr::kable()
```
