---
title: "Pooled Bayesian estimates"
subtitle: "Source notebook for table 2"
author: "PTCy meta-analysis collaboration"
format:
  html:
    toc: true
    code-fold: show
    df-print: kable
execute:
  echo: true
  warning: false
  message: false
---

Table 2 of the article reports the posterior median odds ratio, its 95% credible
interval, and the between-study standard deviation τ for every outcome and comparison.
Those posteriors come from `brms` fits that take hours to run and are not re-fitted
here. What this notebook does is take the **exported posterior summaries** and format
them into the published table, so that the table and the model output cannot drift
apart.

Inputs, from `03_models/set_b/` of the analysis repository:

- `data/models/Table2_setB.csv` — posterior summaries for the adopted (Set B,
  cohort-deduplicated) analysis
- `data/models/freq_results_setB.csv` — frequentist REML concordance checks

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(knitr)

here <- function(...) file.path(if (basename(getwd()) == "notebooks") ".." else ".", ...)

t2   <- read_csv(here("data", "models", "Table2_setB.csv"),     show_col_types = FALSE)
freq <- read_csv(here("data", "models", "freq_results_setB.csv"), show_col_types = FALSE)

glimpse(t2)
```

## Formatting the published table

Odds ratios are set to two decimal places with the middle dot the journal uses for
decimals, and totals are set with a thin space as the thousands separator.

```{r}
#| label: helpers

middot  <- function(x, d = 2) str_replace(sprintf("%.*f", d, x), fixed("."), "·")
or_cri  <- function(or, lo, hi) sprintf("%s (%s–%s)", middot(or), middot(lo), middot(hi))
thin_n  <- function(n) format(n, big.mark = " ", trim = TRUE)

# slug -> (outcome label, comparison), in the order the article presents them
key <- tribble(
  ~slug,          ~outcome,                      ~comparison,
  "c1_os",        "Overall survival",            "C1",
  "c2_os",        "Overall survival",            "C2",
  "c3_os",        "Overall survival",            "C3",
  "c1_nrm",       "Non-relapse mortality",       "C1",
  "c3_nrm",       "Non-relapse mortality",       "C3",
  "c1_rrm",       "Relapse-related mortality",   "C1",
  "c2_rrm",       "Relapse-related mortality",   "C2",
  "c3_rrm",       "Relapse-related mortality",   "C3",
  "c1_agvhd",     "Acute GVHD II–IV",       "C1",
  "c2_agvhd",     "Acute GVHD II–IV",       "C2",
  "c3_agvhd",     "Acute GVHD II–IV",       "C3",
  "c1_cgvhd_ms",  "Chronic GVHD mod–severe","C1",
  "c2_cgvhd_ms",  "Chronic GVHD mod–severe","C2",
  "c1_cgvhd_any", "Chronic GVHD any",       "C1",
  "c3_cgvhd",     "Chronic GVHD any",       "C3",
  "c1_cmv",       "CMV reactivation",            "C1",
  "c2_cmv",       "CMV reactivation",            "C2",
  "c1_bsi",       "Bloodstream infection",       "C1",
  "c1_ifi_any",   "Invasive fungal infection",   "C1",
  "c1_bk",        "BK virus",                    "C1",
  "c2_bk",        "BK virus",                    "C2",
  "c1_irm",       "Infection-related mortality", "C1",
  "c2_irm",       "Infection-related mortality", "C2")

# GRADE ratings are a judgement recorded during the review, not a model
# output. Values match appendix S11 (reassessed 2026-09-30: all VERY LOW).
grade <- tribble(
  ~slug,         ~model, ~GRADE,
  "c1_os",       "m1",   "Very low",
  "c2_os",       "m1",   "Very low",
  "c1_nrm",      "m1",   "Very low",
  "c1_agvhd",    "m1",   "Very low",
  "c2_agvhd",    "m1",   "Very low",
  "c1_cmv",      "m1",   "Very low",
  "c2_cmv",      "m1",   "Very low",
  "c1_bsi",      "m1",   "Very low",
  "c1_ifi_any",  "m1",   "Very low")
```

```{r}
#| label: tbl-pooled
#| tbl-cap: "Bayesian meta-analysis results by outcome and comparison (Set B, cohort-deduplicated). M1 = primary model adjusted for timepoint. M2 (exploratory steroid adjustment) is reported in appendix S9 only. k = number of paired studies. N = total patients across all arms. τ = between-study SD (posterior median). 95% PI = prediction interval for the PTCy effect in a new study, from the posterior of the random PTCy-effect slope."

t2 |>
  filter(model == "m1") |>   # M2 is appendix-only (appendix S9)
  inner_join(key, by = "slug") |>
  left_join(grade, by = c("slug", "model")) |>
  mutate(
    Model = recode(model, m1 = "M1", m2_steroid = "M2"),
    order_out = match(outcome, unique(key$outcome))) |>
  arrange(order_out, comparison, Model) |>
  transmute(
    Outcome    = outcome,
    Comparison = comparison,
    Model,
    k,
    N          = thin_n(n_total),
    `OR (95% CrI)` = or_cri(or_median, ci_low, ci_high),
    `τ`   = middot(tau),
    `95% PI`   = sprintf("%s–%s", middot(pi_low), middot(pi_high)),
    GRADE      = coalesce(GRADE, "—"),
    Source     = source) |>
  kable(align = "llcrrllll")
```

## Frequentist concordance check

Every Bayesian estimate was re-fitted with restricted maximum likelihood (`metafor`) as
a concordance check. Agreement in direction and rough magnitude — not identical point
estimates — is what this check is for; the Bayesian intervals are wider because they
propagate uncertainty in τ.

```{r}
#| label: tbl-freq
#| tbl-cap: "Bayesian posterior medians beside frequentist REML estimates."

t2 |>
  filter(model == "m1") |>
  inner_join(key, by = "slug") |>
  inner_join(freq, by = "slug", suffix = c("", "_freq")) |>
  transmute(Outcome = outcome, Comparison = comparison, k,
            Bayesian = or_cri(or_median, ci_low, ci_high),
            `Frequentist REML` = or_cri(freq_or, freq_lo, freq_hi),
            `I²` = sprintf("%d%%", round(I2))) |>
  kable()
```

## Publication bias

```{r}
#| label: tbl-pubbias
#| tbl-cap: "Robust Bayesian model-averaging (RoBMA) assessment of publication bias."

f <- here("data", "models", "pub_bias_results.csv")
if (file.exists(f)) read_csv(f, show_col_types = FALSE) |> kable() else
  cat("pub_bias_results.csv not vendored into this project.\n")
```
