---
title: "Study and arm characteristics"
subtitle: "Source notebook for table 1"
author: "PTCy meta-analysis collaboration"
format:
  html:
    toc: true
    code-fold: show
    df-print: kable
execute:
  echo: true
  warning: false
  message: false
---

This notebook rebuilds table 1 of the article directly from the extraction database, so
every count in the published table can be traced to the rows it came from.

Inputs, copied verbatim from `02_extraction/` of the analysis repository
(<https://github.com/Russlewisbo/ptcy_metaanalys>):

| File | Rows | Contents |
|:-----|-----:|:---------|
| `data/studies.csv` | 251 | one row per included publication |
| `data/arms.csv` | 525 | one row per treatment arm |
| `data/rob.csv` | — | ROBINS-I / RoB 2 judgements |

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(readr)
library(knitr)

here <- function(...) file.path(if (basename(getwd()) == "notebooks") ".." else ".", ...)

studies <- read_csv(here("data", "studies.csv"), show_col_types = FALSE)
arms    <- read_csv(here("data", "arms.csv"),    show_col_types = FALSE)
rob     <- read_csv(here("data", "rob.csv"),     show_col_types = FALSE)

comparative <- studies |> filter(study_design != "single_arm_descriptive_excluded")

# Several numeric fields are stored as free text in the extraction database: they hold
# "NR" and per-subgroup strings such as "27 (MSD/MUD); 62.5 (haplo)". Take the leading
# number, and always report how many values could not be parsed rather than silently
# dropping them.
as_num <- function(x) {
  if (is.numeric(x)) return(x)
  suppressWarnings(as.numeric(sub("^[^0-9.-]*(-?[0-9.]+).*$", "\\1", x)))
}

cat(sprintf(
  "studies: %d (comparative %d, single-arm descriptive %d)\narms: %d (PTCy %d, comparator %d)\n",
  nrow(studies), nrow(comparative), nrow(studies) - nrow(comparative),
  nrow(arms), sum(arms$arm_role == "PTCy_arm"), sum(arms$arm_role == "comparator_arm")))
```

## Study-level counts

```{r}
#| label: tbl-study-level
#| tbl-cap: "Study-level characteristics, recomputed from `studies.csv`."

pct <- function(n, d) sprintf("%d (%d%%)", n, round(100 * n / d))

fu <- as_num(comparative$follow_up_median_months)
cat(sprintf("follow-up parsed for %d of %d comparative studies (%d not reported/unparseable)\n",
            sum(!is.na(fu)), length(fu), sum(is.na(fu))))

design <- comparative |> count(study_design)
n_comp <- nrow(comparative)

tibble(
  Characteristic = c(
    "Studies included in the review",
    "  Comparative (≥2 arms)",
    "  Single-arm descriptive",
    "Total patients",
    "Randomised controlled trials",
    "  Retrospective cohort",
    "  Registry analysis",
    "  Prospective cohort",
    "Publication year, median (IQR)",
    "Follow-up, months, median (IQR)"),
  Value = c(
    as.character(nrow(studies)),
    as.character(n_comp),
    as.character(nrow(studies) - n_comp),
    format(sum(arms$n_patients, na.rm = TRUE), big.mark = " "),
    pct(design$n[design$study_design == "RCT"], n_comp),
    pct(design$n[design$study_design == "retrospective_cohort"], n_comp),
    pct(design$n[design$study_design == "registry"], n_comp),
    pct(design$n[design$study_design == "prospective_cohort"], n_comp),
    with(comparative, sprintf("%d (%d–%d)",
         median(pub_year, na.rm = TRUE),
         quantile(pub_year, .25, na.rm = TRUE),
         quantile(pub_year, .75, na.rm = TRUE))),
    sprintf("%.0f (%.0f–%.0f)", median(fu, na.rm = TRUE),
            quantile(fu, .25, na.rm = TRUE), quantile(fu, .75, na.rm = TRUE)))
) |> kable()
```

## Arm-level characteristics

Continuous variables are summarised as the median of the arm-level values; the IQR
therefore describes between-arm variability, not between-patient variability.

```{r}
#| label: tbl-arm-level
#| tbl-cap: "Arm-level characteristics by arm role, recomputed from `arms.csv`."

med_iqr <- function(x, d = 1) {
  x <- as_num(x)
  x <- x[!is.na(x)]
  if (!length(x)) return("—")
  q <- quantile(x, c(.25, .5, .75))
  sprintf("%.*f (%.*f–%.*f)", d, q[2], d, q[1], d, q[3])
}

# a categorical "predominant" column summarised as n (%) within arm role
cat_rows <- function(col, labels) {
  arms |>
    mutate(lvl = .data[[col]]) |>
    count(arm_role, lvl) |>
    group_by(arm_role) |>
    mutate(txt = sprintf("%d (%d%%)", n, round(100 * n / sum(n)))) |>
    ungroup() |>
    select(arm_role, lvl, txt) |>
    pivot_wider(names_from = arm_role, values_from = txt) |>
    filter(lvl %in% names(labels)) |>
    mutate(lvl = unname(labels[lvl])) |>
    rename(Characteristic = lvl)
}

num_row <- function(label, col, d = 1) {
  s <- arms |> group_by(arm_role) |> summarise(v = med_iqr(.data[[col]], d), .groups = "drop")
  tibble(Characteristic = label,
         PTCy_arm       = s$v[s$arm_role == "PTCy_arm"],
         comparator_arm = s$v[s$arm_role == "comparator_arm"])
}

donor <- c(haplo = "  Haploidentical", mud = "  MUD (10/10)", mmud = "  MMUD (9/10 or lower)",
           msd = "  MSD", ucb = "  UCB")
graft <- c(pbsc = "  PBSC", bm = "  BM", ucb = "  UCB")
cond  <- c(mac = "  MAC", ric = "  RIC", nma = "  NMA")

bind_rows(
  num_row("Patients per arm, median (IQR)", "n_patients", 0),
  num_row("Age, years, median (IQR)",       "age_median_years"),
  num_row("Male, %",                        "pct_male"),
  tibble(Characteristic = "Donor type (predominant)", PTCy_arm = "", comparator_arm = ""),
  cat_rows("donor_predominant", donor),
  tibble(Characteristic = "Graft source (predominant)", PTCy_arm = "", comparator_arm = ""),
  cat_rows("graft_predominant", graft),
  tibble(Characteristic = "Conditioning (predominant)", PTCy_arm = "", comparator_arm = ""),
  cat_rows("conditioning_predominant", cond),
  num_row("CMV R+, %",           "cmv_recipient_pos_pct"),
  num_row("Steroid exposure, %", "systemic_steroid_for_gvhd_exposure_pct")
) |>
  replace_na(list(PTCy_arm = "—", comparator_arm = "—")) |>
  rename(`PTCy arms` = PTCy_arm, `Comparator arms` = comparator_arm) |>
  kable(align = "lrr")
```

## Risk of bias

```{r}
#| label: tbl-rob
#| tbl-cap: "Overall risk-of-bias judgements."

rob_col <- intersect(c("overall_judgement", "overall", "rob_overall"), names(rob))[1]
if (!is.na(rob_col)) {
  rob |>
    count(Judgement = .data[[rob_col]]) |>
    mutate(`n (%)` = sprintf("%d (%d%%)", n, round(100 * n / sum(n)))) |>
    select(-n) |>
    kable()
} else {
  cat("Columns present in rob.csv:", paste(names(rob), collapse = ", "), "\n")
}
```
